---
title: "Wrangling Exercise"
author: "CMSC320"
date: "February 18, 2016"
output: html_document
---

Use `dplyr` on the `nycflights13` dataset to perform the following data preparation and summarization tasks. In each case print the first 5 lines of the resulting table. 

Recall that to load this dataset:

```{r, eval=FALSE}
library(dplyr)
library(nycflights13)
```

*Question 1*: Filter flights to include only flights with an arrival delay greater than 2 hours (delays are recorded in minutes).

```{r}
# enter your solution here
```

*Question 2*: Select columns in the flights table that contain delay times (note that the function `matches` can be used to select columns with names that partially match a given string. See `?dplyr::select`)

```{r}
# enter solution here
```

*Question 3*: Find the minimum arrival delay (this can be a negative number) in the flights table. Use `summarize`.

```{r}
# enter solution here
```

*Question 4*: Find minimum arrival delay (again, can be a negative number) in the flights table for each _destination_ airport. Use `group_by` and `summarize`.

```{r}
# enter solution here
```

*Question 5*: List the name of **all** airlines and the number of flights for each airline in flights table. The `airlines` table contains airline names, so you will have to perform a join operation.

```{r}
# enter solution here
```
*Question 6*: 

(a) Create a new column (use `mutate`) with total delay time for each flight. 

(b) Replace any missing _total delay times_ with the average (`mean`) total delay. 

(c) Same as (b) but now replace any missing _total delay times_ with the average (`mean`) total delay for the flight's route (i.e., origin/destination combination)

```{r}
# enter solutions here
```

