---
title: "baseball"
output: html_document
---

Look at dollars per win and dollars per run, for each team across time.

Redo the team by payroll table.

- sum salaries by year


```{r}
library(dplyr)
library(Lahman)
data(Salaries)

salaries <- tbl_df(Salaries)
payroll <- salaries %>%
  group_by(teamID, yearID) %>%
  summarize(payroll=sum(salary))
```

Now join to get number of games won, runs scored and rank and keep only data from 1995 on
```{r}
data(Teams)
teams <- tbl_df(Teams)
team_tab <- payroll %>%
  left_join(teams, by=c("teamID", "yearID")) %>%
  select(teamID, franchID, yearID, payroll, wins=W, runs_scored=R, runs_allowed=RA, rank=Rank) %>%
  filter(yearID >= 1995)
```

Plot payroll vs. wins for the years 2000, 2005 and 2010

```{r}
library(ggplot2)
team_tab %>% filter(yearID==2000) %>%
  ggplot(aes(x=payroll, y=wins, label=teamID)) +
    geom_text() +
    geom_smooth() 
```

```{r}
team_tab %>% filter(yearID==2005) %>%
  ggplot(aes(x=payroll, y=wins, label=teamID)) +
    geom_text() +
    geom_smooth() 
```

```{r}
team_tab %>% filter(yearID==2010) %>%
  ggplot(aes(x=payroll, y=wins, label=teamID)) +
    geom_text() +
    geom_smooth() 
```

Calculate dollars per win, and per run

```{r}
team_tab2 <- team_tab %>% 
  mutate(dollars_per_win = payroll / wins) %>%
  mutate(dollars_per_run = payroll / runs_scored) %>%
  mutate(win_expectation = (runs_scored^2) / (runs_scored^2 + runs_allowed^2)) %>%
  mutate(dollars_per_expected_win = payroll / (win_expectation * 162))
```

Now plot dollars per win for each year, for each team

```{r}
team_tab2 %>%
  ggplot(aes(x=yearID, y=dollars_per_win)) +
  geom_line(aes(group=franchID))
```

Now plot dollars per expected win for each year, for each team

```{r}
team_tab2 %>%
  ggplot(aes(x=yearID, y=dollars_per_expected_win)) +
  geom_line(aes(group=franchID))
```

Now center dollars per expected win for each year

```{r}
dew_tab <- team_tab2 %>%
  group_by(yearID) %>%
  summarize(median_DEW=median(dollars_per_expected_win))

team_dew_tab <- team_tab2 %>%
  left_join(dew_tab, by="yearID") %>%
  mutate(relative_dew = dollars_per_expected_win - median_DEW) %>%
  mutate(expected_wins = win_expectation * 162) %>%
  select(franchID, yearID, relative_dew, expected_wins) 
```
plot relative dew
```{r}
team_dew_tab %>%
  filter(teamID %in% c("OAK", "KCA", "NYN", "NYA", "BOS")) %>%
  ggplot(aes(x=yearID, y=relative_dew)) +
  geom_line(aes(group=franchID, color=franchID))  
```

```{r}
team_dew_tab %>%
  filter(teamID %in% c("OAK", "KCA", "NYN", "NYA", "BOS")) %>%
  ggplot(aes(x=yearID, y=relative_dew)) +
  geom_point(aes(color=franchID, size=expected_wins))
```
