---
title: "TidyCode"
output: html_document
---

```{r, echo=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

Workbook for data tidying lecture.
First off: I cloned data for Hadley's paper from [github](https://github.com/hadley/tidyr)

All datasets will be found in this directory afterwards:
```{r}
data_dir <- "tidyr/vignettes"
```

A tidy dataset
----------------

```{r}
library(nycflights13)
head(flights)
```

Headers as values
------------------

```{r}
library(tidyr)
library(dplyr)

pew <- tbl_df(read.csv(file.path(data_dir, "pew.csv"), stringsAsFactors=FALSE, check.names=FALSE))
pew
```

Gather values in columns into a table with `income` and `frequency` columns

```{r}
gather(pew, income, frequency, -religion)
```

Another example: Billboard songs
```{r}
billboard <- tbl_df(read.csv(file.path(data_dir, "billboard.csv"), stringsAsFactors=FALSE))
billboard
```

```
billboard %>% gather(week, rank, wk1:wk76)
```

Multiple variables in one column
--------------------------------------

Tuberculosis case data
```{r}
tb <- tbl_df(read.csv(file.path(data_dir, "tb.csv"), stringsAsFactors = FALSE))
tb
```

```{r}
tb %>% 
  gather(demo, n, -iso2, -year, na.rm=TRUE) %>%
  separate(demo, c("sex", "age"), 1)
```

Multiple types in one table
----------------------------

Some more cleaning up

```{r}
billboard3 <- billboard %>%
  gather(week, rank, wk1:wk76, na.rm=TRUE) %>%
  mutate(
    week = extract_numeric(week),
    date = as.Date(date.entered) + 7 * (week - 1)) %>%
  select(-date.entered)
billboard3
```

Split into song and rank tables

```{r}
song <- billboard3 %>%
  select(artist, track, year, time) %>% 
  unique() %>%
  mutate(song_id = row_number())
song
```

For rank table, we first need the song id so join 
```{r}
# exercise
rank <- billboard3 %>% 
  left_join(song, c("artist", "track", "year", "time")) %>%
  select(c("week","rank","date","song_id"))
rank
```
