---
title: "Regression Project"
author: "Hector Corrada Bravo"
output: html_document
---

```{r, echo=FALSE}
knitr::opts_chunk$set(cache=FALSE)
```

In this part of this project you will practice and experiment with linear regression using
data from [gapminder.org](http://gapminder.org). I recommend spending a little time looking at material there, it is quite an informative site.

We will use a subset of data provided by gapminder provided by [Jennifer Bryan](http://www.stat.ubc.ca/~jenny/) described in it's [github page](https://github.com/jennybc/gapminder).

The following commands load the dataset

```{r}
library(gapminder)
data(gapminder)

head(gapminder)
```

For this exercise you will explore how life expectancy has changed over 50 years across the world, and how economic measures like gross domestic product (GDP) are related to it.

**Exercise 1**: Make a scatter plot of life expectancy across time.

```{r}
library(dplyr)
library(ggplot2)

gapminder %>%
  ggplot(aes(x=year,y=lifeExp)) +
    geom_point() +
  geom_smooth(method=lm)
```

**Question 1**: Is there a general trend (e.g., increasing or decreasing) for life expectancy across time? Is this trend linear? (answering this qualitatively from the plot, you will do a statistical analysis of this question shortly)

_your answer here_

A slightly different way of making the same plot is looking at the distribution of life expectancy across countries as it changes over time:

```{r, fig.width=12}
library(dplyr)
library(ggplot2)

gapminder %>%
  ggplot(aes(x=factor(year), y=lifeExp)) +
    geom_violin() +
    labs(title="Life expectancy over time",
         x = "year",
         y = "life expectancy")
```

This type of plot is called a _violin plot_, and it displays the distribution of the variable in the y-axis for each value of the variable in the x-axis.

**Question 2**: How would you describe the distribution of life expectancy across countries for individual years? Is it skewed, or not? Unimodal or not? Symmetric around it's center?

_your answer here_

Based on this plot, consider the following questions.

**Question 3**: Suppose I fit a linear regression model of life expectancy vs. year (treating it as a continuous variable), and test for an association between year and life expectancy, will you reject the null hypothesis of no association?

_your answer here_

**Question 4**: What would a violin plot of residuals from the linear model in Question 3 vs. year look like?

_your answer here_

**Question 5**: According to the assumptions of the linear regression model, what **should** that violin plot look like?

_your response_

**Exercise 2**: Fit a linear regression model using the `lm` function for life expectancy vs. year (as a continuous variable). Use the `summary` function to look at the resulting model.

```{r}
fit <- lm(lifeExp~year, data=gapminder)
summary(fit)
```

**Question 6**: On average, by how much does life expectancy increase every year around the world?

_your answer here_

**Question 7**: Do you reject the null hypothesis of no association between year and life expectancy? Why?

_your answer here_

**Exercise 3**: Make a violin plot of residuals vs. year for the linear model from Exercise 2.

```{r}
gapminder %>%
  mutate(residuals=residuals(fit)) %>%
  ggplot(aes(x=factor(year), y=residuals)) +
  geom_violin()
```

**Question 8**: Does it match your expectations (Question 4)?

_your answer here_

**Exercise 4**: Make a boxplot (or violin plot) of model residuals vs. continent.

```{r}
gapminder %>%
  mutate(residuals=residuals(fit)) %>%
  ggplot(aes(x=continent, y=residuals)) +
  geom_boxplot()
```

**Question 9**: Is there a dependence between model residual and continent? If so, what would that suggest when performing a regression analysis of life expectancy across time?

_your answer here_

**Exercise 5**: Use `geom_smooth(method=lm)` in ggplot as part of a scatter plot of life expectancy vs. year, grouped by continent (e.g., using the `color` aesthetic mapping).

```{r}
gapminder %>%
  ggplot(aes(x=year,y=lifeExp,color=continent,size=pop)) +
  geom_point() +
  geom_smooth(method=lm)
```

**Question 10**: Based on this plot, should your regression model include an interaction term for continent and year? Why?

_your answer here_

**Exercise 6**: Fit a linear regression model for life expectancy including a term for an interaction between continent and year. Use the `summary` function to show the resulting model.


```{r}
interaction_fit <- lm(lifeExp~year*continent, data=gapminder)
summary(interaction_fit)
```

**Question 11**: Are all estimates in the model significantly different from zero? If not, which are not significantly different from zero?

_your answer here_

**Question 12**: On average, by how much does life expectancy increase each year for each  continent? (You should write code to answer this one)

```{r}
estimates <- coef(interaction_fit)[c(2,7:10)]

round(c(Africa=estimates[1],
  estimates[1]+estimates[-1])*12,2)
```

_your answer here_

**Exercise 7**: You can use the `anova` function to perform an F-test that compares how well two models fit your data. Use it to compare the linear regression models from Exercise 2 (only including year as a covariate) and Exercise 6 (including interaction between year and continent).

```{r}
# your code here
```

**Question 13**: Is the interaction model significantly better than the year-only model? Why?

**Exercise 8**: Make a residuals vs. year violin plot for the interaction model. Comment on how well it matches assumptions of the linear regression model. Do the same for a residuals vs. fitted values model. (Note, you can get fitted values using the `fitted` function on the object returned by `lm`).

```{r}
gapminder %>%
  mutate(residuals=resid(interaction_fit)) %>%
  ggplot(aes(x=factor(year),y=residuals)) +
  geom_violin()
```

_your comments here_
